---
title: "Student Performance Project Summary: Classification"
author: "Applied Statistics Group"
output: word_document
date: "2026-06-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

plot_colors <- c("#2E86AB", "#F18F01", "#C73E1D", "#6A994E", "#7B2CBF", "#00A6A6", "#EF476F")
category_colors <- c("Low" = "#C73E1D", "Medium" = "#F18F01", "High" = "#2E86AB")
accent_color <- "#C1121F"
grid_color <- "gray88"
```

## Project Aim

This project explores university student performance and study habits.  
The main goal is to summarize the dataset compare simple classification methods for predicting GPA-based performance categories.

## Load Data

```{r load-data}
# Change path to run code on your computer
csv_path <- "C:\\Users\\samps\\Documents\\applied_statistic_project\\applied_statistic_project\\global_university_students_performance_habits_10000.csv"

df <- read.csv(csv_path, stringsAsFactors = FALSE, check.names = FALSE)

data.frame(
  Rows = nrow(df),
  Columns = ncol(df)
)
```

## Dataset Overview

```{r dataset-overview}
head(df)
```

## Missing Values and Duplicates

```{r missing-duplicates}
missing_summary <- data.frame(
  Variable = names(df),
  Missing_Values = colSums(is.na(df))
)

missing_summary[missing_summary$Missing_Values > 0, ]

duplicates <- sum(duplicated(df))

data.frame(
  Duplicate_Rows = duplicates
)
```

## Performance Categories

Students were divided into three GPA-based performance groups: Low, Medium, and High.

```{r performance-categories}
df$performance_category <- cut(
  df$GPA,
  breaks = quantile(df$GPA, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE
)

df$performance_category <- factor(
  df$performance_category,
  levels = c("Low", "Medium", "High")
)

table(df$performance_category)
```

```{r performance-category-plot}
barplot(
  table(df$performance_category),
  main = "Performance Category Counts",
  xlab = "Performance Category",
  ylab = "Number of Students",
  col = category_colors[names(table(df$performance_category))],
  border = "white",
  las = 1
)
grid(nx = NA, ny = NULL, col = grid_color)
```

## Data Leakage Controls

The target variable is based on GPA, so GPA must not be used as a predictor. In addition, final exam score and assignment score are direct academic outcome variables, so they are also removed before modeling. Student ID is removed because it is only an identifier.

These leakage-control variables are removed from the numeric dataset used for PCA, nearest-centroid classification, and kNN. They are also removed separately in the decision tree and Naive Bayes sections, because those sections begin again from the full dataset in order to include categorical predictors.


## Numeric Variables Used

```{r numeric-variables}
num_idx <- vapply(df, is.numeric, logical(1))
students.num <- df[, num_idx, drop = FALSE]

leakage_vars <- c(
  "student_id",
  "GPA",
  "final_exam_score",
  "assignment_score"
)

students.num <- students.num[
  ,
  !names(students.num) %in% c(leakage_vars, "university_year"),
  drop = FALSE
]

names(students.num)
```

## Variable Distributions

```{r variable-distributions}
par(mfrow = c(3, 3))

for (var in names(students.num)[1:min(9, length(names(students.num)))]) {
  hist(
    students.num[[var]],
    main = paste("Histogram of", var),
    xlab = var,
    col = plot_colors[(match(var, names(students.num)) - 1) %% length(plot_colors) + 1],
    border = "white"
  )
}

par(mfrow = c(1, 1))
```

## PCA Method

The numeric predictor variables were standardized before PCA. GPA, final exam score, assignment score, and student ID were excluded to avoid leakage or non-informative identifiers. The numeric PCA-based models use the cleaned numeric predictor set created above.

```{r pca-method}
students.scaled <- scale(students.num)

pca_fit <- prcomp(
  students.scaled,
  center = FALSE,
  scale. = FALSE
)

pca_var <- pca_fit$sdev^2
pca_prop <- pca_var / sum(pca_var)
pca_cum <- cumsum(pca_prop)

pca_summary <- data.frame(
  PC = paste0("PC", seq_along(pca_prop)),
  Eigenvalue = round(pca_var, 4),
  Proportion = round(pca_prop, 4),
  Cumulative = round(pca_cum, 4)
)

head(pca_summary, 10)
```

```{r pca-cumulative-plot}
plot(
  pca_cum,
  type = "b",
  pch = 19,
  col = plot_colors[1],
  bg = plot_colors[1],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Cumulative Variance Explained",
  main = "PCA Cumulative Variance Explained"
)
grid(col = grid_color)
```

## Shared Train/Test Split

The same train/test split was used for all classification methods to make model comparisons fair.

```{r train-test-split}
set.seed(123)

n <- nrow(students.scaled)
train_idx <- sample(seq_len(n), size = floor(0.80 * n))
test_idx <- setdiff(seq_len(n), train_idx)

data.frame(
  Training_Rows = length(train_idx),
  Testing_Rows = length(test_idx)
)
```

## Nearest-Centroid Classifier

The nearest-centroid classifier assigns each test student to the performance group whose average position, or centroid, is closest.

This section compares two versions:

1. A nearest-centroid classifier using PCA scores.
2. A nearest-centroid classifier using the original standardized numeric variables.

### Centroid Classifier Functions

```{r nearest-centroid-functions}
classify_nearest_centroid <- function(point, centroids_df, predictor_names) {
  
  distances <- apply(
    centroids_df[, predictor_names, drop = FALSE],
    1,
    function(center) sqrt(sum((point - center)^2))
  )
  
  centroids_df$performance_category[which.min(distances)]
}


evaluate_pc_centroid_classifier <- function(k, pca_fit, performance_category, train_idx, test_idx) {
  
  pc_names <- paste0("PC", 1:k)
  
  pca_k <- as.data.frame(pca_fit$x[, pc_names, drop = FALSE])
  pca_k$performance_category <- performance_category
  
  train_data <- pca_k[train_idx, ]
  test_data  <- pca_k[test_idx, ]
  
  centroid_formula <- as.formula(
    paste("cbind(", paste(pc_names, collapse = ", "), ") ~ performance_category")
  )
  
  centroids <- aggregate(
    centroid_formula,
    data = train_data,
    FUN = mean
  )
  
  test_predictions <- apply(
    test_data[, pc_names, drop = FALSE],
    1,
    classify_nearest_centroid,
    centroids_df = centroids,
    predictor_names = pc_names
  )
  
  test_predictions <- factor(
    test_predictions,
    levels = levels(performance_category)
  )
  
  confusion <- table(
    Actual = test_data$performance_category,
    Predicted = test_predictions
  )
  
  accuracy <- mean(test_data$performance_category == test_predictions)
  
  list(
    k = k,
    accuracy = accuracy,
    confusion = confusion,
    predictions = test_predictions,
    centroids = centroids
  )
}
```

### PCA-Based Centroid Accuracy by Number of PCs

```{r nearest-centroid-pca-results}
max_pcs <- ncol(pca_fit$x)

accuracy_results <- data.frame(
  Number_of_PCs = 1:max_pcs,
  Accuracy = NA_real_,
  Cumulative_Variance = pca_cum[1:max_pcs]
)

for (k in 1:max_pcs) {
  
  result_k <- evaluate_pc_centroid_classifier(
    k = k,
    pca_fit = pca_fit,
    performance_category = df$performance_category,
    train_idx = train_idx,
    test_idx = test_idx
  )
  
  accuracy_results$Accuracy[k] <- result_k$accuracy
}

accuracy_results$Accuracy <- round(accuracy_results$Accuracy, 4)
accuracy_results$Cumulative_Variance <- round(accuracy_results$Cumulative_Variance, 4)

accuracy_results
```

```{r nearest-centroid-pca-plot}
plot(
  accuracy_results$Number_of_PCs,
  accuracy_results$Accuracy,
  type = "b",
  pch = 19,
  col = plot_colors[1],
  bg = plot_colors[1],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy",
  main = "PCA-Based Nearest-Centroid Accuracy",
  ylim = c(0, 1)
)
grid(col = grid_color)

abline(
  h = max(accuracy_results$Accuracy),
  lty = 2,
  col = accent_color,
  lwd = 2
)
```

### PCA-Centroid Elbow Plot

This plot helps choose a practical number of principal components by comparing test accuracy with cumulative variance explained.  

```{r nearest-centroid-pca-elbow-plot}
centroid_best_accuracy_pc <- accuracy_results$Number_of_PCs[
  which.max(accuracy_results$Accuracy)
]

plot(
  accuracy_results$Number_of_PCs,
  accuracy_results$Accuracy,
  type = "b",
  pch = 19,
  lty = 1,
  col = plot_colors[1],
  bg = plot_colors[1],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy / Cumulative Variance",
  main = "PCA-Centroid Elbow Plot",
  ylim = c(0, 1)
)
grid(col = grid_color)

lines(
  accuracy_results$Number_of_PCs,
  accuracy_results$Cumulative_Variance,
  type = "b",
  pch = 17,
  lty = 2,
  col = plot_colors[2],
  bg = plot_colors[2],
  lwd = 2
)

abline(
  v = centroid_best_accuracy_pc,
  lty = 3,
  col = accent_color,
  lwd = 2
)

legend(
  "bottomright",
  legend = c("Accuracy", "Cumulative variance", "Maximum accuracy PC"),
  col = c(plot_colors[1], plot_colors[2], accent_color),
  pch = c(19, 17, NA),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

### Best PCA-Based Centroid Model

The PCA-based centroid classifier was tested using different numbers of principal components.  
This shows whether the model needs many dimensions or whether most of the useful classification information is already captured by the first few principal components.

```{r best-pca-centroid}
best_pc <- accuracy_results$Number_of_PCs[which.max(accuracy_results$Accuracy)]

best_centroid_pca <- evaluate_pc_centroid_classifier(
  k = best_pc,
  pca_fit = pca_fit,
  performance_category = df$performance_category,
  train_idx = train_idx,
  test_idx = test_idx
)

data.frame(
  Best_Number_of_PCs = best_pc,
  Best_PCA_Centroid_Accuracy = round(best_centroid_pca$accuracy, 4),
  Cumulative_Variance = accuracy_results$Cumulative_Variance[best_pc]
)
```

### Confusion Matrix for Best PCA-Based Centroid Model

```{r best-pca-centroid-confusion}
best_centroid_pca$confusion
```

## Nearest-Centroid Classifier Using Original Standardized Variables

This version uses the original standardized numeric variables instead of PCA scores.  
It helps show whether PCA improved or reduced centroid classification accuracy.

```{r original-variable-centroid}
centroid_original_data <- as.data.frame(students.scaled)
centroid_original_data$performance_category <- df$performance_category

train_original <- centroid_original_data[train_idx, ]
test_original  <- centroid_original_data[test_idx, ]

predictor_names <- names(students.num)

centroid_original_formula <- as.formula(
  paste("cbind(", paste(predictor_names, collapse = ", "), ") ~ performance_category")
)

centroids_original <- aggregate(
  centroid_original_formula,
  data = train_original,
  FUN = mean
)

original_centroid_predictions <- apply(
  test_original[, predictor_names, drop = FALSE],
  1,
  classify_nearest_centroid,
  centroids_df = centroids_original,
  predictor_names = predictor_names
)

original_centroid_predictions <- factor(
  original_centroid_predictions,
  levels = levels(df$performance_category)
)

original_centroid_confusion <- table(
  Actual = test_original$performance_category,
  Predicted = original_centroid_predictions
)

original_centroid_accuracy <- mean(
  test_original$performance_category == original_centroid_predictions
)

data.frame(
  Model = "Nearest centroid, original standardized variables",
  Accuracy = round(original_centroid_accuracy, 4)
)
```

### Confusion Matrix for Original-Variable Centroid Model

```{r original-variable-centroid-confusion}
original_centroid_confusion
```

## Centroid Model Comparison

```{r centroid-model-comparison}
centroid_model_comparison <- data.frame(
  Model = c(
    "PCA centroid",
    "Original-variable centroid"
  ),
  Accuracy = c(
    best_centroid_pca$accuracy,
    original_centroid_accuracy
  )
)

centroid_model_comparison$Accuracy <- round(
  centroid_model_comparison$Accuracy,
  4
)

centroid_model_comparison
```

```{r centroid-model-comparison-plot}
old_par <- par(mar = c(8, 5, 4, 2))

barplot(
  centroid_model_comparison$Accuracy,
  names.arg = c("PCA\ncentroid", "Original-variable\ncentroid"),
  ylim = c(0, 1),
  ylab = "Test Accuracy",
  main = "Centroid Classifier Accuracy Comparison",
  las = 1,
  col = plot_colors[1:2],
  border = "white"
)
grid(nx = NA, ny = NULL, col = grid_color)

par(old_par)
```
## PCA Dimension Reduction Result

Although the original-variable centroid model uses all standardized numeric predictors, the PCA-based centroid model may achieve similar accuracy using far fewer dimensions.

This is important because PCA is mainly a dimension-reduction method.  
If one or a few principal components give similar accuracy to the full original-variable model, then PCA has successfully compressed most of the useful classification information into a smaller number of variables.
```{r}
pc1_accuracy <- accuracy_results$Accuracy[accuracy_results$Number_of_PCs == 1]

dimension_reduction_summary <- data.frame(
  Model = c(
    "PCA centroid using 1 PC",
    paste("Best PCA centroid using", best_pc, "PCs"),
    "Original-variable centroid"
  ),
  Dimensions_Used = c(
    1,
    best_pc,
    length(predictor_names)
  ),
  Accuracy = c(
    pc1_accuracy,
    round(best_centroid_pca$accuracy, 4),
    round(original_centroid_accuracy, 4)
  )
)

dimension_reduction_summary
```

### Accuracy Compared with Number of Dimensions
```{r pca-dimensions-accuracy-plot}
old_par <- par(mar = c(5, 5, 4, 10), xpd = TRUE)

plot(
  dimension_reduction_summary$Dimensions_Used,
  dimension_reduction_summary$Accuracy,
  pch = 19,
  cex = 1.5,
  col = plot_colors[seq_len(nrow(dimension_reduction_summary))],
  bg = plot_colors[seq_len(nrow(dimension_reduction_summary))],
  xlab = "Number of Dimensions Used",
  ylab = "Test Accuracy",
  main = "Accuracy Compared with Number of Dimensions",
  ylim = c(0, 1),
  xlim = c(0, max(dimension_reduction_summary$Dimensions_Used) + 5)
)
grid(col = grid_color)

text(
  dimension_reduction_summary$Dimensions_Used,
  dimension_reduction_summary$Accuracy,
  labels = dimension_reduction_summary$Model,
  pos = 4,
  cex = 0.75
)

par(old_par)
```
The centroid results show that PCA can match the original-variable centroid model when all principal components are retained. However, 2 PCs are already very close to having all PCs.

## PC1 vs PC2 Plot with Class Centroids

This plot shows how the performance categories are positioned in the first two principal components.  
The star-shaped points represent the class centroids calculated from the training data.

```{r pc1-pc2-centroid-plot}
pca_plot_data <- as.data.frame(pca_fit$x[, 1:2])
pca_plot_data$performance_category <- df$performance_category

train_pca_plot <- pca_plot_data[train_idx, ]

centroids_pc12 <- aggregate(
  cbind(PC1, PC2) ~ performance_category,
  data = train_pca_plot,
  FUN = mean
)

cloud_colors <- c(
  "Low" = "red",
  "Medium" = "orange",
  "High" = "blue"
)

centroid_color <- "black"

plot(
  pca_plot_data$PC1,
  pca_plot_data$PC2,
  col = adjustcolor(cloud_colors[pca_plot_data$performance_category], alpha.f = 0.45),
  pch = 16,
  xlab = "PC1",
  ylab = "PC2",
  main = "Students and Class Centroids on PC1 and PC2"
)
grid(col = grid_color)

points(
  centroids_pc12$PC1,
  centroids_pc12$PC2,
  pch = 8,
  col = centroid_color,
  cex = 2.8,
  lwd = 3
)

legend(
  "topright",
  legend = levels(pca_plot_data$performance_category),
  col = cloud_colors[levels(pca_plot_data$performance_category)],
  pch = 16,
  title = "Performance Category",
  bty = "n"
)

legend(
  "bottomright",
  legend = "Centroids",
  col = centroid_color,
  pch = 8,
  pt.cex = 2,
  lwd = 3,
  bty = "n"
)
```

## kNN Classifier

The kNN classifier predicts each test student's performance category based on the most similar students in the training set.  
Because kNN is distance-based, the model was tested in two ways:

1. kNN using the original standardized numeric variables.
2. kNN using PCA scores.

This allows us to compare whether PCA can reduce the number of dimensions while preserving or improving classification accuracy.

```{r knn-setup}
library(class)

predictor_names <- names(students.num)

knn_data <- as.data.frame(students.scaled)
knn_data$performance_category <- df$performance_category

train_data <- knn_data[train_idx, ]
test_data  <- knn_data[test_idx, ]

k_values <- c(1, 3, 5, 7, 9, 11, 15, 21, 31, 51, 75, 101)
```

## Original-Variable kNN Tuning

This version uses all standardized numeric predictor variables.

```{r original-knn-tuning}
original_knn_tuning_results <- data.frame(
  k = k_values,
  Accuracy = NA_real_
)

for (i in seq_along(k_values)) {

  k_val <- k_values[i]

  pred <- knn(
    train = train_data[, predictor_names],
    test = test_data[, predictor_names],
    cl = train_data$performance_category,
    k = k_val
  )

  original_knn_tuning_results$Accuracy[i] <- mean(
    test_data$performance_category == pred
  )
}

original_knn_tuning_results$Accuracy <- round(
  original_knn_tuning_results$Accuracy,
  4
)

original_knn_tuning_results
```

```{r original-knn-tuning-plot}
plot(
  original_knn_tuning_results$k,
  original_knn_tuning_results$Accuracy,
  type = "b",
  pch = 19,
  col = plot_colors[3],
  bg = plot_colors[3],
  lwd = 2,
  xlab = "Number of Neighbors, k",
  ylab = "Test Accuracy",
  main = "Original-Variable kNN Accuracy by k",
  ylim = c(0, 1)
)
grid(col = grid_color)
```

```{r best-original-knn}
best_original_k <- original_knn_tuning_results$k[
  which.max(original_knn_tuning_results$Accuracy)
]

best_original_knn_accuracy <- max(
  original_knn_tuning_results$Accuracy
)

best_original_knn_predictions <- knn(
  train = train_data[, predictor_names],
  test = test_data[, predictor_names],
  cl = train_data$performance_category,
  k = best_original_k
)

best_original_knn_confusion <- table(
  Actual = test_data$performance_category,
  Predicted = best_original_knn_predictions
)

data.frame(
  Model = "Original-variable kNN",
  Best_k = best_original_k,
  Dimensions_Used = length(predictor_names),
  Accuracy = round(best_original_knn_accuracy, 4)
)
```

### Confusion Matrix for Best Original-Variable kNN

```{r best-original-knn-confusion}
best_original_knn_confusion
```

## PCA-Based kNN Tuning

This version uses PCA scores instead of the original standardized variables.  
Both the number of principal components and the number of neighbors are tuned.

```{r pca-knn-tuning}
pca_scores <- as.data.frame(pca_fit$x)

pc_values <- 1:ncol(pca_scores)

pca_knn_tuning_results <- expand.grid(
  Number_of_PCs = pc_values,
  k = k_values
)

pca_knn_tuning_results$Accuracy <- NA_real_

for (i in seq_len(nrow(pca_knn_tuning_results))) {

  pc_count <- pca_knn_tuning_results$Number_of_PCs[i]
  k_val <- pca_knn_tuning_results$k[i]

  pc_names <- paste0("PC", 1:pc_count)

  train_pca <- pca_scores[train_idx, pc_names, drop = FALSE]
  test_pca  <- pca_scores[test_idx, pc_names, drop = FALSE]

  pred <- knn(
    train = train_pca,
    test = test_pca,
    cl = df$performance_category[train_idx],
    k = k_val
  )

  pca_knn_tuning_results$Accuracy[i] <- mean(
    df$performance_category[test_idx] == pred
  )
}

pca_knn_tuning_results$Accuracy <- round(
  pca_knn_tuning_results$Accuracy,
  4
)

pca_knn_tuning_results
```

```{r pca-knn-tuning-plot}
plot(
  pca_knn_tuning_results$Number_of_PCs,
  pca_knn_tuning_results$Accuracy,
  pch = 19,
  col = adjustcolor(plot_colors[4], alpha.f = 0.6),
  bg = adjustcolor(plot_colors[4], alpha.f = 0.6),
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy",
  main = "PCA-kNN Accuracy by Number of PCs",
  ylim = c(0, 1)
)
grid(col = grid_color)

legend(
  "bottomright",
  legend = "Each point represents one k value",
  pch = 19,
  bty = "n"
)
```

### PCA-kNN Elbow Plot

Because PCA-kNN is tuned using both the number of PCs and the number of neighbors, this summary keeps the best kNN accuracy for each PC count.  
This makes it easier to choose a practical number of PCs by looking for where accuracy begins to level off.

```{r pca-knn-pc-summary}
pca_knn_pc_summary <- aggregate(
  Accuracy ~ Number_of_PCs,
  data = pca_knn_tuning_results,
  FUN = max
)

names(pca_knn_pc_summary)[names(pca_knn_pc_summary) == "Accuracy"] <- "Best_Accuracy"

pca_knn_best_k_by_pc <- do.call(
  rbind,
  lapply(
    split(pca_knn_tuning_results, pca_knn_tuning_results$Number_of_PCs),
    function(x) x[which.max(x$Accuracy), c("Number_of_PCs", "k")]
  )
)

pca_knn_pc_summary <- merge(
  pca_knn_pc_summary,
  pca_knn_best_k_by_pc,
  by = "Number_of_PCs"
)

pca_knn_pc_summary$Cumulative_Variance <- pca_cum[
  pca_knn_pc_summary$Number_of_PCs
]

pca_knn_pc_summary$Best_Accuracy <- round(
  pca_knn_pc_summary$Best_Accuracy,
  4
)

pca_knn_pc_summary$Cumulative_Variance <- round(
  pca_knn_pc_summary$Cumulative_Variance,
  4
)

pca_knn_pc_summary
```

```{r pca-knn-elbow-plot}
pca_knn_best_accuracy_pc <- pca_knn_pc_summary$Number_of_PCs[
  which.max(pca_knn_pc_summary$Best_Accuracy)
]

plot(
  pca_knn_pc_summary$Number_of_PCs,
  pca_knn_pc_summary$Best_Accuracy,
  type = "b",
  pch = 19,
  lty = 1,
  col = plot_colors[4],
  bg = plot_colors[4],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Best Test Accuracy / Cumulative Variance",
  main = "PCA-kNN Elbow Plot",
  ylim = c(0, 1)
)
grid(col = grid_color)

lines(
  pca_knn_pc_summary$Number_of_PCs,
  pca_knn_pc_summary$Cumulative_Variance,
  type = "b",
  pch = 17,
  lty = 2,
  col = plot_colors[2],
  bg = plot_colors[2],
  lwd = 2
)

abline(
  v = pca_knn_best_accuracy_pc,
  lty = 3,
  col = accent_color,
  lwd = 2
)

legend(
  "bottomright",
  legend = c("Best accuracy across k", "Cumulative variance", "Maximum accuracy PC"),
  col = c(plot_colors[4], plot_colors[2], accent_color),
  pch = c(19, 17, NA),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

```{r best-pca-knn}
best_pca_knn_row <- pca_knn_tuning_results[
  which.max(pca_knn_tuning_results$Accuracy),
]

best_pca_knn_pcs <- best_pca_knn_row$Number_of_PCs
best_pca_knn_k <- best_pca_knn_row$k
best_pca_knn_accuracy <- best_pca_knn_row$Accuracy

best_pca_knn_pc_names <- paste0("PC", 1:best_pca_knn_pcs)

best_pca_knn_predictions <- knn(
  train = pca_scores[train_idx, best_pca_knn_pc_names, drop = FALSE],
  test = pca_scores[test_idx, best_pca_knn_pc_names, drop = FALSE],
  cl = df$performance_category[train_idx],
  k = best_pca_knn_k
)

best_pca_knn_confusion <- table(
  Actual = df$performance_category[test_idx],
  Predicted = best_pca_knn_predictions
)

data.frame(
  Model = "PCA-based kNN",
  Best_k = best_pca_knn_k,
  Best_Number_of_PCs = best_pca_knn_pcs,
  Accuracy = round(best_pca_knn_accuracy, 4)
)
```

### Confusion Matrix for Best PCA-Based kNN

```{r best-pca-knn-confusion}
best_pca_knn_confusion
```

## kNN Model Comparison

```{r knn-model-comparison}
knn_model_comparison <- data.frame(
  Model = c(
    "Original-variable kNN",
    "PCA-based kNN"
  ),
  Best_k = c(
    best_original_k,
    best_pca_knn_k
  ),
  Dimensions_Used = c(
    length(predictor_names),
    best_pca_knn_pcs
  ),
  Accuracy = c(
    best_original_knn_accuracy,
    best_pca_knn_accuracy
  )
)

knn_model_comparison$Accuracy <- round(
  knn_model_comparison$Accuracy,
  4
)

knn_model_comparison
```

```{r knn-model-comparison-plot}
old_par <- par(mar = c(8, 5, 4, 2))

barplot(
  knn_model_comparison$Accuracy,
  names.arg = c("Original-variable\nkNN", "PCA-based\nkNN"),
  ylim = c(0, 1),
  ylab = "Test Accuracy",
  main = "kNN Accuracy Comparison",
  las = 1,
  col = plot_colors[3:4],
  border = "white"
)
grid(nx = NA, ny = NULL, col = grid_color)

par(old_par)
```

## kNN Dimension Reduction Result

The PCA-based kNN model can be compared with the original-variable kNN model in terms of both accuracy and number of dimensions used.

```{r knn-dimension-reduction-summary}
knn_dimension_summary <- data.frame(
  Model = c(
    "Original-variable kNN",
    "PCA-based kNN"
  ),
  Dimensions_Used = c(
    length(predictor_names),
    best_pca_knn_pcs
  ),
  Accuracy = c(
    best_original_knn_accuracy,
    best_pca_knn_accuracy
  )
)

knn_dimension_summary$Accuracy <- round(
  knn_dimension_summary$Accuracy,
  4
)

knn_dimension_summary
```

```{r knn-dimensions-accuracy-plot}
old_par <- par(mar = c(5, 5, 4, 10), xpd = TRUE)

plot(
  knn_dimension_summary$Dimensions_Used,
  knn_dimension_summary$Accuracy,
  pch = 19,
  cex = 1.5,
  col = plot_colors[3:4],
  bg = plot_colors[3:4],
  xlab = "Number of Dimensions Used",
  ylab = "Test Accuracy",
  main = "kNN Accuracy Compared with Number of Dimensions",
  ylim = c(0, 1),
  xlim = c(0, max(knn_dimension_summary$Dimensions_Used) + 5)
)
grid(col = grid_color)

text(
  knn_dimension_summary$Dimensions_Used,
  knn_dimension_summary$Accuracy,
  labels = knn_dimension_summary$Model,
  pos = 4,
  cex = 0.75
)

par(old_par)
```

The kNN results show whether PCA can reduce the predictor space while preserving classification accuracy.  
If PCA-based kNN performs similarly to original-variable kNN using fewer dimensions, then PCA provides a simpler distance-based classification model.


## Decision Tree Classifier

A decision tree was added because it can use both numeric and categorical predictors.  
This allows the analysis to include variables that were not used in centroid classification or kNN.

GPA, final exam score, assignment score, and student ID were removed from the predictors. GPA was removed because it was used to create the performance category, while final exam score and assignment score were removed because they are direct academic outcome variables. Student ID was removed because it is only an identifier.

```{r decision-tree-setup}
library(rpart)

tree_data <- df

tree_data$performance_category <- factor(
  tree_data$performance_category,
  levels = c("Low", "Medium", "High")
)

leakage_vars <- c(
  "student_id",
  "GPA",
  "final_exam_score",
  "assignment_score"
)

tree_predictor_names <- setdiff(
  names(tree_data),
  c(leakage_vars, "performance_category")
)

is_categorical <- vapply(
  tree_data[, tree_predictor_names, drop = FALSE],
  function(x) is.character(x) || is.factor(x),
  logical(1)
)

categorical_predictors <- tree_predictor_names[is_categorical]

categorical_level_counts <- sapply(
  tree_data[, categorical_predictors, drop = FALSE],
  function(x) length(unique(x))
)

high_cardinality_vars <- names(
  categorical_level_counts[categorical_level_counts > 10]
)

tree_predictor_names <- setdiff(
  tree_predictor_names,
  high_cardinality_vars
)

character_columns <- vapply(
  tree_data[, tree_predictor_names, drop = FALSE],
  is.character,
  logical(1)
)

tree_data[, tree_predictor_names][character_columns] <- lapply(
  tree_data[, tree_predictor_names][character_columns],
  factor
)

tree_model_data <- tree_data[
  ,
  c(tree_predictor_names, "performance_category"),
  drop = FALSE
]

tree_model_data$row_id <- seq_len(nrow(tree_model_data))

tree_model_data <- tree_model_data[
  complete.cases(tree_model_data[, setdiff(names(tree_model_data), "row_id")]),
]

train_tree <- tree_model_data[tree_model_data$row_id %in% train_idx, ]
test_tree  <- tree_model_data[tree_model_data$row_id %in% test_idx, ]

train_tree$row_id <- NULL
test_tree$row_id <- NULL

data.frame(
  Predictors_Used = length(tree_predictor_names),
  Removed_High_Cardinality_Categorical_Variables = paste(high_cardinality_vars, collapse = ", ")
)
```

```{r decision-tree-model}
tree_fit <- rpart(
  performance_category ~ .,
  data = train_tree,
  method = "class",
  control = rpart.control(
    cp = 0.001,
    minsplit = 100,
    minbucket = 50,
    maxdepth = 4,
    xval = 0,
    maxcompete = 2,
    maxsurrogate = 0,
    usesurrogate = 0
  )
)

tree_predictions <- predict(
  tree_fit,
  newdata = test_tree,
  type = "class"
)

tree_confusion <- table(
  Actual = test_tree$performance_category,
  Predicted = tree_predictions
)

tree_accuracy <- mean(
  test_tree$performance_category == tree_predictions
)

data.frame(
  Model = "Decision tree",
  Predictors_Used = "Numeric and low-cardinality categorical variables",
  Accuracy = round(tree_accuracy, 4)
)
```

### Decision Tree Plot

```{r decision-tree-plot, fig.width=9, fig.height=6}
if (requireNamespace("rpart.plot", quietly = TRUE)) {

  rpart.plot::rpart.plot(
    tree_fit,
    type = 2,
    extra = 104,
    fallen.leaves = TRUE,
    main = "Decision Tree for Performance Category",
    box.palette = "GnBu",
    shadow.col = "gray80",
    branch.col = "gray50"
  )

} else {

  plot(
    tree_fit,
    uniform = TRUE,
    margin = 0.1,
    main = "Decision Tree for Performance Category",
    box.palette = "GnBu",
    shadow.col = "gray80",
    branch.col = "gray50"
  )

  text(
    tree_fit,
    use.n = TRUE,
    all = TRUE,
    cex = 0.7
  )
}
```

### Decision Tree Variable Importance

```{r decision-tree-variable-importance}
tree_importance <- data.frame(
  Variable = names(tree_fit$variable.importance),
  Importance = as.numeric(tree_fit$variable.importance)
)

tree_importance <- tree_importance[
  order(tree_importance$Importance, decreasing = TRUE),
]

head(tree_importance, 10)
```

```{r decision-tree-variable-importance-plot}
top_tree_importance <- head(tree_importance, 10)

old_par <- par(mar = c(5, 12, 4, 2))

barplot(
  rev(top_tree_importance$Importance),
  names.arg = rev(top_tree_importance$Variable),
  horiz = TRUE,
  las = 1,
  col = rev(rep(plot_colors, length.out = nrow(top_tree_importance))),
  border = "white",
  xlab = "Variable Importance",
  main = "Top Decision Tree Predictors"
)
grid(nx = NULL, ny = NA, col = grid_color)

par(old_par)
```

The decision tree provides an interpretable model that can use both numeric and categorical predictors.  
This makes it useful as a comparison against the PCA-based and distance-based methods, which only used numeric predictors.


## Bayesian Classifier

A Bayesian classifier was added as another comparison model.  
This model uses Bayes' theorem to predict the most likely performance category for each student.

The Naive Bayes classifier assumes that predictors are conditionally independent within each class.  
Although this assumption is often simplified, Naive Bayes can still perform well and provides a useful comparison against centroid, kNN, and decision tree classifiers.

GPA, final exam score, assignment score, and student ID were removed to avoid data leakage.

```{r bayesian-classifier-setup}
if (!requireNamespace("e1071", quietly = TRUE)) {
  install.packages("e1071")
}

library(e1071)

bayes_data <- df

bayes_data$performance_category <- factor(
  bayes_data$performance_category,
  levels = c("Low", "Medium", "High")
)

leakage_vars <- c(
  "student_id",
  "GPA",
  "final_exam_score",
  "assignment_score"
)

bayes_predictor_names <- setdiff(
  names(bayes_data),
  c(leakage_vars, "performance_category")
)

is_categorical <- vapply(
  bayes_data[, bayes_predictor_names, drop = FALSE],
  function(x) is.character(x) || is.factor(x),
  logical(1)
)

categorical_predictors <- bayes_predictor_names[is_categorical]

categorical_level_counts <- sapply(
  bayes_data[, categorical_predictors, drop = FALSE],
  function(x) length(unique(x))
)

high_cardinality_vars_bayes <- names(
  categorical_level_counts[categorical_level_counts > 10]
)

bayes_predictor_names <- setdiff(
  bayes_predictor_names,
  high_cardinality_vars_bayes
)

character_columns <- vapply(
  bayes_data[, bayes_predictor_names, drop = FALSE],
  is.character,
  logical(1)
)

bayes_data[, bayes_predictor_names][character_columns] <- lapply(
  bayes_data[, bayes_predictor_names][character_columns],
  factor
)

bayes_model_data <- bayes_data[
  ,
  c(bayes_predictor_names, "performance_category"),
  drop = FALSE
]

bayes_model_data$row_id <- seq_len(nrow(bayes_model_data))

bayes_model_data <- bayes_model_data[
  complete.cases(bayes_model_data[, setdiff(names(bayes_model_data), "row_id")]),
]

train_bayes <- bayes_model_data[bayes_model_data$row_id %in% train_idx, ]
test_bayes  <- bayes_model_data[bayes_model_data$row_id %in% test_idx, ]

train_bayes$row_id <- NULL
test_bayes$row_id <- NULL

data.frame(
  Predictors_Used = length(bayes_predictor_names),
  Removed_High_Cardinality_Categorical_Variables = paste(high_cardinality_vars_bayes, collapse = ", ")
)
```

```{r bayesian-classifier-model}
bayes_fit <- naiveBayes(
  performance_category ~ .,
  data = train_bayes
)

bayes_predictions <- predict(
  bayes_fit,
  newdata = test_bayes,
  type = "class"
)

bayes_confusion <- table(
  Actual = test_bayes$performance_category,
  Predicted = bayes_predictions
)

bayes_accuracy <- mean(
  test_bayes$performance_category == bayes_predictions
)

data.frame(
  Model = "Naive Bayes classifier",
  Predictors_Used = "Numeric and low-cardinality categorical variables",
  Accuracy = round(bayes_accuracy, 4)
)
```

### Confusion Matrix for Bayesian Classifier

```{r bayesian-classifier-confusion}
bayes_confusion
```

The Bayesian classifier provides another interpretable comparison model.  
It uses probability-based classification and can include both numeric and categorical predictors, making it useful to compare with the decision tree.


## PCA-Based Bayesian Classifier

A PCA-based Naive Bayes classifier was also tested.  
This version uses principal component scores instead of the original predictor variables.

The purpose is to check whether PCA can reduce the number of dimensions while still preserving enough information for probability-based classification.

```{r pca-bayesian-classifier-tuning}
pca_bayes_tuning_results <- data.frame(
  Number_of_PCs = 1:ncol(pca_scores),
  Accuracy = NA_real_,
  Cumulative_Variance = pca_cum[1:ncol(pca_scores)]
)

for (pc_count in 1:ncol(pca_scores)) {
  
  pc_names <- paste0("PC", 1:pc_count)
  
  train_pca_bayes <- data.frame(
    pca_scores[train_idx, pc_names, drop = FALSE],
    performance_category = df$performance_category[train_idx]
  )
  
  test_pca_bayes <- data.frame(
    pca_scores[test_idx, pc_names, drop = FALSE],
    performance_category = df$performance_category[test_idx]
  )
  
  pca_bayes_fit <- naiveBayes(
    performance_category ~ .,
    data = train_pca_bayes
  )
  
  pca_bayes_predictions <- predict(
    pca_bayes_fit,
    newdata = test_pca_bayes,
    type = "class"
  )
  
  pca_bayes_tuning_results$Accuracy[pc_count] <- mean(
    test_pca_bayes$performance_category == pca_bayes_predictions
  )
}

pca_bayes_tuning_results$Accuracy <- round(
  pca_bayes_tuning_results$Accuracy,
  4
)

pca_bayes_tuning_results$Cumulative_Variance <- round(
  pca_bayes_tuning_results$Cumulative_Variance,
  4
)

pca_bayes_tuning_results
```

```{r pca-bayesian-classifier-plot}
plot(
  pca_bayes_tuning_results$Number_of_PCs,
  pca_bayes_tuning_results$Accuracy,
  type = "b",
  pch = 19,
  col = plot_colors[5],
  bg = plot_colors[5],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy",
  main = "PCA-Based Naive Bayes Accuracy",
  ylim = c(0, 1)
)
grid(col = grid_color)

abline(
  h = max(pca_bayes_tuning_results$Accuracy),
  lty = 2,
  col = accent_color,
  lwd = 2
)
```

### PCA-Naive Bayes Elbow Plot

This plot helps choose a practical number of PCs for the PCA-based Bayesian classifier.  

```{r pca-bayesian-classifier-elbow-plot}
pca_bayes_best_accuracy_pc <- pca_bayes_tuning_results$Number_of_PCs[
  which.max(pca_bayes_tuning_results$Accuracy)
]

plot(
  pca_bayes_tuning_results$Number_of_PCs,
  pca_bayes_tuning_results$Accuracy,
  type = "b",
  pch = 19,
  lty = 1,
  col = plot_colors[5],
  bg = plot_colors[5],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy / Cumulative Variance",
  main = "PCA-Naive Bayes Elbow Plot",
  ylim = c(0, 1)
)
grid(col = grid_color)

lines(
  pca_bayes_tuning_results$Number_of_PCs,
  pca_bayes_tuning_results$Cumulative_Variance,
  type = "b",
  pch = 17,
  lty = 2,
  col = plot_colors[2],
  bg = plot_colors[2],
  lwd = 2
)

abline(
  v = pca_bayes_best_accuracy_pc,
  lty = 3,
  col = accent_color,
  lwd = 2
)

legend(
  "bottomright",
  legend = c("Accuracy", "Cumulative variance", "Maximum accuracy PC"),
  col = c(plot_colors[1], plot_colors[2], accent_color),
  pch = c(19, 17, NA),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 2),
  bty = "n"
)
```

```{r best-pca-bayesian-classifier}
best_pca_bayes_pc <- pca_bayes_tuning_results$Number_of_PCs[
  which.max(pca_bayes_tuning_results$Accuracy)
]

best_pca_bayes_pc_names <- paste0("PC", 1:best_pca_bayes_pc)

best_pca_bayes_train <- data.frame(
  pca_scores[train_idx, best_pca_bayes_pc_names, drop = FALSE],
  performance_category = df$performance_category[train_idx]
)

best_pca_bayes_test <- data.frame(
  pca_scores[test_idx, best_pca_bayes_pc_names, drop = FALSE],
  performance_category = df$performance_category[test_idx]
)

best_pca_bayes_fit <- naiveBayes(
  performance_category ~ .,
  data = best_pca_bayes_train
)

best_pca_bayes_predictions <- predict(
  best_pca_bayes_fit,
  newdata = best_pca_bayes_test,
  type = "class"
)

best_pca_bayes_confusion <- table(
  Actual = best_pca_bayes_test$performance_category,
  Predicted = best_pca_bayes_predictions
)

best_pca_bayes_accuracy <- mean(
  best_pca_bayes_test$performance_category == best_pca_bayes_predictions
)

data.frame(
  Model = "PCA-based Naive Bayes",
  Best_Number_of_PCs = best_pca_bayes_pc,
  Accuracy = round(best_pca_bayes_accuracy, 4),
  Cumulative_Variance = pca_bayes_tuning_results$Cumulative_Variance[best_pca_bayes_pc]
)
```

### Confusion Matrix for Best PCA-Based Bayesian Classifier

```{r best-pca-bayesian-classifier-confusion}
best_pca_bayes_confusion
```

## Bayesian Classifier Comparison

```{r bayesian-classifier-comparison}
bayes_model_comparison <- data.frame(
  Model = c(
    "Original-variable Naive Bayes",
    "PCA-based Naive Bayes"
  ),
  Predictors_Used = c(
    "Numeric and categorical variables",
    paste(best_pca_bayes_pc, "principal component(s)")
  ),
  Accuracy = c(
    bayes_accuracy,
    best_pca_bayes_accuracy
  )
)

bayes_model_comparison$Accuracy <- round(
  bayes_model_comparison$Accuracy,
  4
)

bayes_model_comparison
```

```{r bayesian-classifier-comparison-plot}
old_par <- par(mar = c(8, 5, 4, 2))

barplot(
  bayes_model_comparison$Accuracy,
  names.arg = c("Original-variable
Naive Bayes", "PCA-based
Naive Bayes"),
  ylim = c(0, 1),
  ylab = "Test Accuracy",
  main = "Naive Bayes Accuracy Comparison",
  las = 1,
  col = plot_colors[5:6],
  border = "white"
)
grid(nx = NA, ny = NULL, col = grid_color)

par(old_par)
```

The PCA-based Bayesian classifier shows whether dimensionality reduction can preserve classification accuracy for a probability-based model.  
If the PCA-based version performs similarly to the original-variable Naive Bayes classifier while using fewer dimensions, then PCA provides a simpler Bayesian classification model.


## PCA Accuracy by Number of Principal Components

This plot compares the PCA-based classifiers directly. For kNN, the line uses the best test accuracy across the tested neighbor values for each number of principal components, so there is only one kNN line.

```{r pca-accuracy-by-pcs-combined-plot}
combined_pc_accuracy <- data.frame(
  Number_of_PCs = accuracy_results$Number_of_PCs,
  Centroid = accuracy_results$Accuracy,
  kNN = pca_knn_pc_summary$Best_Accuracy[
    match(accuracy_results$Number_of_PCs, pca_knn_pc_summary$Number_of_PCs)
  ],
  Naive_Bayes = pca_bayes_tuning_results$Accuracy[
    match(accuracy_results$Number_of_PCs, pca_bayes_tuning_results$Number_of_PCs)
  ]
)

plot(
  combined_pc_accuracy$Number_of_PCs,
  combined_pc_accuracy$Centroid,
  type = "b",
  pch = 19,
  lty = 1,
  col = plot_colors[1],
  bg = plot_colors[1],
  lwd = 2,
  xlab = "Number of Principal Components",
  ylab = "Test Accuracy",
  main = "PCA-Based Classifier Accuracy by Number of PCs",
  ylim = c(0, 1)
)
grid(col = grid_color)

lines(
  combined_pc_accuracy$Number_of_PCs,
  combined_pc_accuracy$kNN,
  type = "b",
  pch = 17,
  lty = 2,
  col = plot_colors[4],
  bg = plot_colors[4],
  lwd = 2
)

lines(
  combined_pc_accuracy$Number_of_PCs,
  combined_pc_accuracy$Naive_Bayes,
  type = "b",
  pch = 15,
  lty = 3,
  col = plot_colors[5],
  bg = plot_colors[5],
  lwd = 2
)

legend(
  "bottomright",
  legend = c(
    "Nearest centroid",
    "kNN, best k for each PC count",
    "Naive Bayes"
  ),
  col = c(plot_colors[1], plot_colors[4], plot_colors[5]),
  pch = c(19, 17, 15),
  lty = c(1, 2, 3),
  lwd = 2,
  bty = "n"
)
```

## Final Classifier Comparison

```{r final-classifier-comparison}
final_model_comparison <- data.frame(
  Model = c(
    "PCA centroid",
    "Original-variable centroid",
    "Original-variable kNN",
    "PCA-based kNN",
    "Decision tree",
    "Original-variable Naive Bayes",
    "PCA-based Naive Bayes"
  ),
  Predictors_Used = c(
    paste(best_pc, "principal component(s)"),
    "Original standardized numeric variables",
    "Original standardized numeric variables",
    paste(best_pca_knn_pcs, "principal component(s)"),
    "Numeric and categorical variables",
    "Numeric and categorical variables",
    paste(best_pca_bayes_pc, "principal component(s)")
  ),
  Accuracy = c(
    best_centroid_pca$accuracy,
    original_centroid_accuracy,
    best_original_knn_accuracy,
    best_pca_knn_accuracy,
    tree_accuracy,
    bayes_accuracy,
    best_pca_bayes_accuracy
  )
)

final_model_comparison$Accuracy <- round(
  final_model_comparison$Accuracy,
  4
)

final_model_comparison
```

```{r final-classifier-comparison-plot}
old_par <- par(mar = c(9, 5, 4, 2))

barplot(
  final_model_comparison$Accuracy,
  names.arg = c(
    "PCA\ncentroid",
    "Original\ncentroid",
    "Original\nkNN",
    "PCA\nkNN",
    "Decision\ntree",
    "Original\nBayes",
    "PCA\nBayes"
  ),
  ylim = c(0, 1),
  ylab = "Test Accuracy",
  main = "Final Classifier Accuracy Comparison",
  las = 1,
  col = plot_colors,
  border = "white"
)
grid(nx = NA, ny = NULL, col = grid_color)

par(old_par)
```


## Final Conclusions

This project compared several simple classifiers for predicting GPA-based student performance categories. The same 80/20 train/test split was used throughout, which makes the model comparisons fair. Leakage was controlled by removing GPA, final exam score, assignment score, and student ID from the predictor sets.

The best overall test accuracy was achieved by the PCA-based Naive Bayes classifier. However, this model used 13 principal components, so its advantage is mainly predictive accuracy rather than dimensionality reduction.

The best balance between accuracy and simplicity came from the PCA-based kNN model. It achieved accuracy very close to the best model while using only 4 principal components instead of the full 13 original numeric dimensions. This suggests that PCA was most useful for kNN, where reducing the distance-based feature space can remove noise and preserve the most useful structure.

The nearest-centroid model achieved the same accuracy with PCA scores and with the original standardized numeric variables when all 13 dimensions were used. This means PCA did not improve the centroid classifier, but it also did not reduce performance when the full PCA representation was retained.

The decision tree was not the most accurate model, but it was the most interpretable. Its variable importance results showed that class attendance percentage and study hours per day were the strongest predictors, followed by mental stress level. These variables provide the clearest practical explanation of the performance categories.

Overall, PCA was useful, but its benefit depended on the classifier. PCA-based kNN gave the strongest dimensionality-reduction result, PCA-based Naive Bayes gave the highest accuracy, and the decision tree gave the clearest explanation of which student habits mattered most.

```{r final-written-summary}
data.frame(
  Conclusion = c(
    "Highest accuracy",
    "Best accuracy-simplicity tradeoff",
    "Most interpretable model",
    "Most important tree predictor",
    "Main methodological control"
  ),
  Finding = c(
    paste0("PCA-based Naive Bayes, accuracy = ", round(best_pca_bayes_accuracy, 4)),
    paste0("PCA-based kNN, ", best_pca_knn_pcs, " PCs, accuracy = ", round(best_pca_knn_accuracy, 4)),
    paste0("Decision tree, accuracy = ", round(tree_accuracy, 4)),
    as.character(tree_importance$Variable[1]),
    "Removed GPA, final exam score, assignment score, and student ID from predictors"
  )
)
```

## Key Points for Poster

- GPA was converted into three balanced performance categories: Low, Medium, and High.
- Numeric predictor variables were standardized before PCA and kNN.
- GPA was excluded from predictors to avoid using the target variable as an input.
- The same 80/20 train/test split was used for all classification models.
- PCA was used to summarize variation in the numeric predictors.
- The nearest-centroid classifier was tested using different numbers of principal components.
- The PCA-based centroid model was compared with a centroid model using the original standardized variables.
- Classification performance was evaluated using test-set accuracy and confusion matrices.
- kNN was tested using both original standardized variables and PCA scores.
- PCA-based kNN was tuned using both the number of neighbors and the number of principal components..
- PCA compressed the numeric predictors into a smaller number of dimensions while preserving similar classification accuracy.
- The first few principal components captured enough structure for some models, especially PCA-based kNN, but the strongest PCA centroid and PCA Naive Bayes results used all 13 PCs.
- A decision tree was added to include both numeric and categorical predictors.
- The decision tree provides an interpretable comparison model with a visual tree structure.
- A Naive Bayes classifier was added as a probability-based Bayesian comparison model.
- 2 Principal Components were almost just as good as more!!! In the poster we can compare accuracy between just 2 PCs, and the "optimal" (most accurate) number of PCs
